---
title: "Challenge: Hey Birders, don't look up!"
author: Ryan Hull
date: "2025-07-11"
format: html
toc: true
---

This challenge was made by Camila Diaz, Christophe Brabant, David Hunt, Katherine Hébert, Laura Pollock, Lindsay Trottier, Lucas Eckert, and Lukas Van Riel of the Living Data Project working group.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_knit$set(root.dir = here::here())

library(arrow)
library(dbplyr, warn.conflicts = FALSE)
```

## About the challenge
This challenge targets the birders of Canada, inviting them to observe 
less-sampled organisms during their iNaturalist outings.

Being the only Blitz the Gap challenge that is members-only, it is also a more
direct measure of Blitz the Gap's outreach success than any other challenge.

## What this script does
Tallies the total observation count broken down by iconic species group.
Presents average observations per user.
Compares past vs present %observations which are birds to observe the trend.

## Load data
```{r Data}

# All inaturalist observations for Canada until Jan 2025
inat_pq <- arrow::open_dataset("C:/Users/Dell/OneDrive - McGill University/Laura's Lab_Group - Blitz the Gap/iNaturalist Canada parquet/iNat_non_sensitive_data_Jan2025.parquet")

# Observations from the birders challenge
birders_pq <- arrow:open_datset(....)

# Breaking down our challenge's observations
obs <- inat_pq |>
  # filter to mammals in this case
  filter(iconic_taxon_name %in% "Aves") |>
  # select columns we want to keep
  select(c(observed_on_string,
           iconic_taxon_name, 
           scientific_name, 
           coordinates_obscured,
           place_county_name)) |>
  collect()