---
title: "Trailblazers"
author: Leeya Nagpal
date: "2025-07-04"
output: html_document
---

This analysis is based on the challenge script `Accessible-parks.R` here: [https://github.com/CIEE-Living-Data-Project/blitz-the-gap/tree/main/02_scripts/Trailblazers](https://github.com/CIEE-Living-Data-Project/blitz-the-gap/tree/main/02_scripts/Trailblazers).

This challenge was made by Lukas Van Riel as part of the Living Data Project working group.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '~/Documents/GitHub/BTG-analyse-the-gap/')
library(ggplot2)
library(dplyr)
library(arrow)
library(dbplyr, warn.conflicts = FALSE)
library(duckdb)
library(terra)
library(mapview) # for interactive maps
library(ineq)
library(leafpop)
```

## What this script does:
This script uses the code from the original Trailblazers Challenge (designed by Lukas Van Riel) to visualize the new priority taxa post-Blitz the Gap. It additionally compares the old and new data, highlighting the areas that have changed priority after the duration of Blitz the Gap.


## Data

```{r Data}

parks <- sf::st_read(dsn = "00_rawdata/trailblazers/national_parks/national parks.shp")
canada <- sf::read_sf("00_rawdata/trailblazers/canada-polygon/canada.outline.shp")
base.5k <- terra::rast("00_rawdata/base-layers/canada.base.5k.tiff")
base.1k <- terra::rast("00_rawdata/base-layers/canada.base.1k.tiff")

#inat parquet
inat_pq <- arrow::open_dataset("~/Desktop/RProjects/Blitz-the-gap/00_rawdata/trailblazers/iNat/iNat_non_sensitive_data_Jan2025.parquet")

# Create query and collect afterwards
query <- inat_pq |> 
  filter(quality_grade == "research") |> #filter to research grade
  filter(positional_accuracy < 1000) |> #filter to accurate 
  select(iconic_taxon_name,scientific_name,longitude,latitude) |>
  collect()
old <- na.omit(query)

new <- read.csv("newParquet.csv")

# Rework
inat.v <- terra::vect(new, geom = c("longitude","latitude"), crs="+proj=longlat +ellps=WGS84")
inat.v <- terra::project(inat.v, crs(base.1k))
inat.v <- terra::crop(inat.v, canada)

# simplify parks polygons
parks.simple = sf::st_simplify(parks, preserveTopology = FALSE, dTolerance = 1000) # 1 km tolerance
parks.sv <- vect(parks.simple)


```

## Analysis

Analysis required for trailblazers:

```{r Analysis}
inat.pr <- project(inat.v, crs(parks.sv))


obs_with_polygons <- terra::intersect(inat.pr, parks.sv)
obs_counts <- as.data.frame(table(obs_with_polygons$NAME_E))

p1k <- obs_counts %>% 
  filter(Freq > 1000) %>% 
  arrange(desc(Freq)) %>% 
  rename(Park_name=Var1, N_obs=Freq) %>% 
  write.csv("~/Desktop/RProjects/Blitz-the-gap/00_rawdata/trailblazers/parks-observations/1k.csv")

##### Gini index:
pk <- pk1[,-1]

new_gini_park <- function(park.name){
  print(park.name) # To keep track
  
  # Check which ones are done already
  parks.done <- read.csv("01_outdata/trailblazers/Gini/parks-gini.csv")[,-1]
  if(park.name %in% parks.done$Park_name){
    return()
  } else {
  park <- parks %>% 
    sf::st_as_sf() %>% 
    filter(NAME_E == park.name) %>% 
    sf::st_simplify(preserveTopology = FALSE, dTolerance = 2000) %>% #simplify outline to 2k
    vect() %>% 
    project(crs(inat.v))
  
  inat.park <- terra::crop(inat.v, park)
  inat.rast <- rasterize(inat.park, base.1k, fun = "length")
  
  r_masked <- mask(inat.rast, park)
  r_crop <- crop(r_masked, park)  # tighten extent to polygon
  
  # Extract values inside polygon
  vals <- values(r_crop, na.rm = TRUE)
  
  # Compute Gini index
  gini <- Gini(vals)
  
  output <- c(park.name,
              pk$N_obs[pk$Park_name == park.name],
              round(gini, 2),
              ifelse(gini > 0.9, 
                                      1,
                                      ifelse(gini > 0.75, 2, 3)))
  
  write.csv(rbind(parks.done, output), "01_outdata/trailblazers/Gini/new-parks-gini.csv")
  }
}
 
sapply(pk$Park_name, gini_park)

#COMPARE WITH OLD MAP
oldGini <- read.csv("parks_gini.csv")[,1]
newGini <- read.csv("new-parks-gini.csv")[,1]

colnames(oldGini)[colnames(oldGini) == "priority"] <- "priority_pre"
colnames(newGini)[colnames(newGini) == "priority"] <- "priority_post"

compare <- full_join(pre, post, by = "Park_name")
compare <- compare %>%
  mutate(priority_changed = priority_pre != priority_post,
         gini_changed = round(as.numeric(gini.x), 2) != round(as.numeric(gini.y), 2))

priority_changes <- compare %>%
  filter(priority_changed | gini_changed) %>%
  arrange(desc(gini_changed))

```

## Results

Plots Below:
1) Park priority map after Blitz the Gap
2) Comparison of park priority pre and post BtG
```{r Results, echo = FALSE}

if (file.exists("01_outdata/trailblazers/gini_parks_map.rds")){
  gini_parks_map <- readRDS("01_outdata/trailblazers/new_gini_parks_map.rds")
} else{

## Create objects to plot
park_gini <- read.csv("01_outdata/trailblazers/Gini/parks-gini.csv")[,-1] %>% 
  select(-priority)

gini_parks_map <- sf::st_as_sf(parks) %>%
  mutate(Park_name = gsub("\\(Recreational Class\\)|\\(Waterway Class\\)|\\(Natural Environment Class\\)|\\(Natural Environment Class\\)|\\(Québec\\)",
                          "", NAME_E) %>% trimws()) %>% 
  select(Park_name, geometry) %>% 
  right_join(park_gini, by="Park_name") %>% 
  filter(gini >= 0) %>% 
  mutate(Priority = factor(ifelse(gini > 0.85, 
                            1,
                            ifelse(gini > 0.75, 2, 3)))) %>% 
  rename(Park = Park_name)
  
saveRDS(gini_parks_map, "01_outdata/trailblazers/new_gini_parks_map.rds")
}

parks_centroid <- new_gini_parks_map %>% 
  sf::st_centroid()

colors <- c("#fc8d59", "#ffffbf", "#99d594")


## NEW MAP VISUAL
mapviewOptions(basemaps = c("OpenStreetMap"),
               na.color = "transparent")

m = mapview(new_gini_parks_map, 
            basemaps = "OpenStreetMap",
            legend = T,
            layer.name = list("Priority"),
            zcol = "Priority",
            col.regions = colors,
            label = paste0("<b>", new_gini_parks_map$Park,"</b><br>"),
            popup = popupTable(new_gini_parks_map, zcol = c("Park", "Priority"))) +
  mapview(parks_centroid,
          legend = F,
          zcol = "Priority",
          col.regions = colors,
          label = paste0("<b>", new_gini_parks_map$Park,"</b><br>"),
          popup = popupTable(new_gini_parks_map, zcol = c("Park", "Priority")))

#change map
mapview(priority_changes, 
            basemaps = "OpenStreetMap",
            legend = T,
            layer.name = list("Priority"),
            zcol = "priority_changed",
            col.regions = colors)



```
