---
title: "Challenge: Make a Splash"
author: Leeya Nagpal
date: "2025-07-08"
format: html
toc: true
---

This analysis is based on the challenge script `fw.R` here: [https://github.com/CIEE-Living-Data-Project/blitz-the-gap/tree/main/02_scripts/freshwater_challenge](https://github.com/CIEE-Living-Data-Project/blitz-the-gap/tree/main/02_scripts/freshwater_challenge).

This challenge was made by Lindsay Trottier and Christophe Brabant as part of the Living Data Project working group.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '~/Documents/GitHub/BTG-analyse-the-gap/')

library(arrow)
library(dbplyr, warn.conflicts = FALSE)
library(duckdb)
library(terra)

```

## What this script does

This script uses the code from the original Make a Splash Challenge (designed by Lindsay Trottier and Christophe Brabant) to visualize the new priority areas post-Blitz the Gap. It additionally compares the old and new data, highlighting the areas that have changed priority taxa after the duration of Blitz the Gap.

## Load Data
```{r Data}

inat_pq <- arrow::open_dataset("C:/Users/dogpa/OneDrive - McGill University/Laura's Lab_Group - inat-canada/iNat_non_sensitive_data_Jan2025.parquet")

query <- inat_pq |> 
  # Summarise number of obs per species, per species group
  group_by(iconic_taxon_name, scientific_name) |>
  summarize(total_obs = n()) |> 
  # load the query into our R session
  collect()

## omit nas
old<-na.omit(query)

new <- read.csv("C:/Users/dogpa/SpeciesList/relevantSpecies/june1-18.csv")

#species
canada_gbif_clean <- read.csv("C:/Users/linds/OneDrive - McGill University/LivingData_Blitz/can_gbif_names.csv")

canada = sf::read_sf("canada.outline.shp")
base.50k = terra::rast("canada.base.5k.tiff")
 
boat = read_sf("canada_launches_all.gpkg")
endemism = read_sf("endemism_all.gpkg")
lakes = read_sf("lhy_000c16a_e.shp")
coast = read_sf("coast2.gpkg")


```

## Analysis

```{r Analysis}

inat.v = terra::vect(obs, geom = c("longitude","latitude"),
                     crs="+proj=longlat +ellps=WGS84")
inat.v = terra::project(inat.v, crs(base.50k))
inat.v = terra::crop(inat.v, canada)
fw.density = rasterize(inat.v, base.50k, fun = "count")

#Project all other data on "canada" crs
endemism <- st_transform(endemism, crs = st_crs(canada))
boat <- st_transform(boat, crs = st_crs(canada))
lakes <- st_transform(lakes, crs = st_crs(canada)) 
coast <- st_transform(coast, crs = st_crs(canada))

coast2 = st_intersection(coast, canada)
coast2 = coast2 %>% select(c(HYDROUID, NAME, PRUID, geom))

water = rbind(lakes, coast2 %>% rename(geometry=geom))

#Get count of boats in each feature
intersections <- st_intersects(water, boat) 
boat_counts <- lengths(intersections)  
water$boat_count <- boat_counts
rm(boat_counts, intersections)

#Get mean endemism per feature
idx <- st_intersects(endemism, water)
endemism_sub <- endemism[lengths(idx) > 0, ]

lake_endemism <- st_intersection(water, endemism_sub)
lake_endemism <- lake_endemism %>%
  st_drop_geometry() %>%
  group_by(HYDROUID) %>%
  summarize(mean_endemism = mean(total_endemism, na.rm = TRUE))

lake_canada_endemism = left_join(water, lake_endemism, by="HYDROUID")

# Get observation per feature

crs(fw.density) <- st_crs(st_crs(canada))$wkt # a WKT string

obs = terra::extract(fw.density, lake_canada_endemism)

obs <-  obs %>% 
  group_by(ID) %>% 
  summarize(observation = sum(count, na.rm = TRUE))

lake_canada_endemism <- lake_canada_endemism %>%
  mutate(ID := seq_len(nrow(.))) %>%
  left_join(obs, by="ID")

new_lake_final = lake_canada_endemism %>% select(c(HYDROUID, NAME, boat_count,
                                               mean_endemism, observation))

write_sf(new_lake_final, "new_fw_challenge.gpkg")


summary(final$mean_endemism)
quantile(final$mean_endemism, na.rm=T, seq(0.90,1))

summary(final$observation)
quantile(final$observation, na.rm=T, seq(0.9,1))

final$priority = NA
final$priority[final$mean_endemism >=  0.3549 & final$observation <= 5] <- 1
final$priority[final$mean_endemism >=  0.3549 & final$observation > 5] <- 2
final$priority[final$mean_endemism <  0.3549 & final$observation <= 5] <-3
final$priority[final$mean_endemism <  0.3549 & final$observation > 5] <- 4
final$priority = as.factor(final$priority)
levels(final$priority)

final = final %>% filter(!is.na(Priority))
final =final %>% select(c(NAME, boat_count, geom, priority)) %>%
  rename('Boat Access' = boat_count, 'Name' = NAME, 'Priority' = priority) 
final <- st_transform(final, crs="EPSG:4326")
st_write(final, "new_water2.geojson")

final2 = final %>% filter(!is.na(priority))
final3 = final2 %>% select(c(NAME, boat_count, geom, priority)) %>%
  rename('Boat Access' = boat_count, 'Name' = NAME, 'Priority' = priority) 

write_sf(final3, "new_data_for_challenge_map.gpkg")

final3 = read_sf("new_data_for_challenge_map.gpkg")

final3 <- st_transform(final3, crs="EPSG:4326")

st_write(final3, "new_water.geojson")


```

## Results

Plots Below:
1) Priority map after Blitz the Gap
2) Comparison of priority areas pre and post BtG
```{r Results, echo = FALSE}

pal = c("#D36135", "#7FB069", "#277CCB")
m=mapview(final3, 
        col.regions = pal,
        zcol = "Priority", 
        lwd=0,
        na.color = "transparent") 

```

