---
title: "Challenge: MayBAs"
author: Leeya Nagpal
date: "2025-07-09"
format: html
toc: true
---

This analysis is based on the challenge script `prioritise-mayBAs.R` here: [https://github.com/CIEE-Living-Data-Project/blitz-the-gap/tree/main/02_scripts/mayBA](https://github.com/CIEE-Living-Data-Project/blitz-the-gap/tree/main/02_scripts/mayBA).

This challenge was made by Katherine Hebert  as part of the Living Data Project working group.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '~/Documents/GitHub/BTG-analyse-the-gap/')

library(sf)
library(raster)
library(mapview)
library(gdalcubes)
library(rstac)
library(stars)
library(terra)
library(ggplot2)
library(leaflet)
library(leafpop)

```

## What this script does

This script uses the code from the original MayBA Challenge (designed by Katherine Hebert) to visualize the new priority areas post-Blitz the Gap. It additionally compares the old and new data, highlighting the areas that have changed priority taxa after the duration of Blitz the Gap.

## Load Data
```{r Data}

inat_pq <- arrow::open_dataset("C:/Users/dogpa/OneDrive - McGill University/Laura's Lab_Group - inat-canada/iNat_non_sensitive_data_Jan2025.parquet")

query <- inat_pq |> 
  # Summarise number of obs per species, per species group
  group_by(iconic_taxon_name, scientific_name) |>
  summarize(total_obs = n()) |> 
  # load the query into our R session
  collect()

## omit nas
old<-na.omit(query)

new <- read.csv("C:/Users/dogpa/SpeciesList/relevantSpecies/june1-18.csv")


```

## Analysis

```{r Analysis}

```

## Results

Plots Below:
1) Priority map after Blitz the Gap
2) Comparison of priority areas pre and post BtG
```{r Results, echo = FALSE}

```

