---
title: "Challenge: the other 99%"
author: Leeya Nagpal 
format: html
date: "2025-07-08"
toc: true
---

This analysis is based on the challenge script `99-percent.R` here: [https://github.com/CIEE-Living-Data-Project/blitz-the-gap/tree/main/02_scripts/99-percent](https://github.com/CIEE-Living-Data-Project/blitz-the-gap/tree/main/02_scripts/99-percent).

This challenge was made by Lucas Eckert as part of the Living Data Project working group.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '~/Documents/GitHub/BTG-analyse-the-gap/')

library(ggplot2)
library(dplyr)
library(arrow)
library(dbplyr, warn.conflicts = FALSE)
library(duckdb)
library(terra)
library(mapview)
library(sf)
library(rmapshaper)
library(leafpop)

```

## What this script does

This script uses the code from the original 99% Challenge (designed by Lucas Eckert) to visualize the new priority areas post-Blitz the Gap. It additionally compares the old and new data, highlighting the areas that have changed priority taxa after the duration of Blitz the Gap.

## Load Data
```{r Data}

inat_pq <- arrow::open_dataset("C:/Users/dogpa/OneDrive - McGill University/Laura's Lab_Group - inat-canada/iNat_non_sensitive_data_Jan2025.parquet")

query <- inat_pq |> 
  # Summarise number of obs per species, per species group
  group_by(iconic_taxon_name, scientific_name) |>
  summarize(total_obs = n()) |> 
  # load the query into our R session
  collect()

## omit nas
old<-na.omit(query)

new <- read.csv("C:/Users/dogpa/SpeciesList/relevantSpecies/june1-18.csv")


## counties
counties<-terra::vect("counties_of_canada.gpkg")

```

## Analysis

Generating the 99% Map with new BtG data, and comparing differences to the old data

```{r Analysis}

if (){## inat layer
inat_vect<-terra::vect(new, geom = c("longitude","latitude"), crs="+proj=longlat +ellps=WGS84")
inat_vect<-terra::project(inat_vect, crs(counties))

## obs
obs_counts<-terra::extract(counties, inat_vect)

country.data<-data.frame("County"=counties$CDUID)
country.data$n.points=NA

for (i in 1:nrow(country.data)){
  c=country.data$County[i]
  poly=counties[which(counties$CDUID==c)]
  points<-crop(inat_vect,ext(poly))
  country.data$n.points[i]=length(which(terra::relate(poly,points,relation="intersects")))
  remove(c,poly,points)
  print(i)
}

country.data<-mutate(country.data, CDUID=County)

county_df<-as.data.frame(values(counties)) %>% left_join(country.data) %>% 
  mutate(obs_density=n.points/LANDAREA) %>% select(-County)

top_counties<-county_df[order(county_df$obs_density, decreasing = T),]
top_counties$cum_area<-cumsum(top_counties$LANDAREA)/sum(top_counties$LANDAREA)
top_counties$cum_obs<-cumsum(top_counties$n.points)/sum(top_counties$n.points)

top_counties[100:111,]

saveRDS(top_counties, file = "new-county-data.RDS")
}

values(counties)<-left_join(values(counties), select(top_counties, CDUID, obs_density))
counties$obs_density_log<-log10(counties$obs_density)

counties$priority<-NA
counties$priority[counties$obs_density>35.5]<-"low"
counties$priority[counties$obs_density<35.5 & counties$obs_density>9.55]<-"medium"
counties$priority[counties$obs_density<9.55]<-"high"
counties$priority<-factor(counties$priority, levels = c("high","medium","low"))


## COMPARE OLD AND NEW
old_99 <- readRDS("county-data.RDS")

pre_df <- as.data.frame(values(old_99)) %>%
  select(CDUID, priority_pre = priority)

post_df <- as.data.frame(values(top_counties)) %>%
  select(CDUID, priority_post = priority)

comparison <- full_join(old_99, top_counties, by = "CDUID") %>%
  mutate(priority_changed = priority_pre != priority_post)

# View changes
changes <- comparison %>% filter(priority_changed)

# Summary of changes
table(comparison$priority_pre, comparison$priority_post, useNA = "ifany")

# How many changed
sum(comparison$priority_changed, na.rm = TRUE)

```

## Results

Plots Below:
1) Priority map after Blitz the Gap
2) Comparison of priority areas pre and post BtG
```{r Results, echo = FALSE}
map_data<-counties
values(map_data)<-values(map_data) %>% select(census_district=CDNAME, observations_per_km2=obs_density, priority) %>%
  mutate(observations_per_km2=ceiling(observations_per_km2))

map_data_sf<-st_as_sf(map_data)
map_data_simp<-st_simplify(map_data_sf, dTolerance = 1000, preserveTopology = TRUE)
map_data_simp2<-ms_simplify(map_data_simp, keep=0.05, keep_shapes = F)

map99<-mapview(map_data_simp2,
        zcol="priority",
        col.regions=c("darkgreen","gold","darkred"),
        layer.name="Priority",
        label = map_data_simp2$census_district,
        popup = popupTable(
          st_drop_geometry(map_data_simp2)[, c("census_district", "observations_per_km2", "priority")],
          feature.id = F,
          row.numbers = F))


# Reattach to current counties for mapping
new_map <- counties
values(new_map) <- left_join(
  values(new_map),
  comparison,
  by = "CDUID"
)

mapview(counties, zcol = "priority_changed", col.regions = c("gray", "red"))
```

